Tracing the Hercules stream around the Galaxy

### 2010-09-12 tex/response.txt

Dear Editor,

Please find below the response to the referees' comments for the
submission "Tracing the Hercules Stream around the Galaxy" by myself
(MS# ApJ80617).

I have made changes following the referee's suggestions, in particular
adding a general discussion to section 5 which discusses the kind of
tracers needed and the number of stars needed to be observed to detect
the Hercules feature. I also added two panels to figure 5 to address
some of the referee's comments. All other changes were minor and in
response to the referee.

@@

	 The paper titled Tracing the Hercules stream around the Galaxy by Jo
	 Bovy is concerned with the origin and detectability of this
	 elusive stream. Overall the paper is well written and
	 clear. I have a few major concerns and some minor concerns
	 that should be addressed before the paper can be accepted for
	 publication.

	 Major concerns:

	 1. Sect 4, page 6, last paragraph: The author discusses in
	 some detail, and shows plots to illustrate, the possible
	 implications of e.g.  errors in the distance determination
	 etc. The one error that is not address explicitly in the text
	 or plots is the error on the measured radial velocities - how
	 well do we need to determine these velocities in order to be
	 able to unambiguously detect the Hercules stream? 1, 5, or 10
	 km/s? Or some other number? And what happens when V0 is
	 absolutely known or if it also has an error? Perhaps the
	 result is that the errors do no matter, but in that case this
	 should be shown and arguments to why it does not matter
	 should be given. Also, will either the RVS on board Gaia or
	 APOGEE deliver radial velocities with small enough errors to
	 detect the Hercules stream?

I added a panel to Figure 5 showing the influence of realistic
line-of-sight velocity uncertainties. The expected uncertainties for
APOGEE and HERMES are below 1 km/s; for such small uncertainties the
uncertainty-free prediction and the uncertainty-convolved prediction
are indistinguishable (which is clear from the new panel).

@@

	 2. A brief discussion about which tracers, e.g. dwarf stars,
	 red clump, red giant stars, to be used to test the model
	 should also be included. For example, is it necessary to
	 reach as far away from the solar neighbourhood as possible
	 (eg giants), or is the major test to get as high volume
	 density of tracers as possible (eg cool dwarfs). Such a
	 discussion would be highly useful for anyone wanting to
	 construct surveys for the very purpose of detecting
	 Hercules. This discussion does not need to be long.

This discussion is part of the new "general" discussion at the start
of section 5.

	 3. I find Discussion in Sect 5 (for both Gaia and APOGEE)
	 somewhat sketchy. For example, no mention of the fact that
	 Gaia will provide radial velocities for stars down to roughly
	 17 magnitude is mentioned.  Should the reader conclude that
	 Hercules could be detected with the tangential velocities
	 alone? In that case, how does that work with the tests
	 presented in this article? Again, here the intrinsic
	 brightness of the tracers is of some interest.

I added a reference to the line-of-sight velocity accuracies expected
for Gaia, and an explicit comment that the velocities of giants will
be measured accurately enough to verify the predictions of this
paper. While the Hercules feature could certainly be detected in
tangential velocities alone, Gaia will have plenty of line-of-sight
velocities to make this prediction unnecessary.

### 2010-11-10 tex/hercules.tex
hercules: proof changes

In \figurename~\ref{fig:rphi1d} I show the predicted line-of-sight
velocity distributions in various locations near the Solar circle all
around the Galaxy, obtained by integrating the full two-dimensional
velocity distribution over the velocity component perpendicular to the
line of sight. This \figurename\ is no longer symmetric with respect
to shifts of 180$^{\circ}$ degree in azimuth, since the line-of-sight
direction breaks this symmetry. The gray curve is the line-of-sight
distribution we would observe if the targeted stars were in a steady
state in an axisymmetric potential with a Dehnen distribution function
(\eqnname~[\ref{eq:fdehnen}]). Both line-of-sight velocity
distributions are normalized to have equal mass since I assume that
our knowledge of the Galaxy and the tracer population will not be good
enough that we can test predicted differences in the absolute number
of stars at a given location between the axisymmetric and the
bar-resonance predictions. These differences are also largely a
function of the particular form and parameters of the distribution
function that we choose to represent the stellar disk before bar
formation.

@@

To further analyze the directions to target to best see the Hercules
feature in the line-of-sight velocities, I have calculated the
line-of-sight velocity distribution on a 100 by 20 grid in
Galactocentric azimuth and radius. As a measure of the difference
between the bar-resonance model and the steady-state, axisymmetric
model, I have calculated the Kullback-Leibler divergence between these
predicted distributions \citep[\eg,][]{mackay}
\begin{equation}
D_{\text{KL}}(P||Q) = \int \dd x\,p(x)\,\ln \frac{p(x)}{q(x)}\,,
\end{equation}
where, in this case, $x$ is the line-of-sight velocity, $p(x)$ is the
predicted distribution for the bar-origin model of the Hercules
stream, and $q(x)$ is the axisymmetric prediction. The
Kullback-Leibler divergence is only one among many possible choices to
quantify the difference between two probability distributions. It can
be interpreted as the expected weight of evidence for the prediction
by the bar-resonance model over the fully axisymmetric prediction that
can be expected from each sampled star. As such, locations with the
maximal Kullback-Leibler divergence are those in which the Hercules
feature stands out most prominently in the line-of-sight.

@@

Finally, in \figurename~\ref{fig:1dvar} I show the influence of
various nuisance and dynamical parameters on the predicted
line-of-sight velocity distribution for the most promising region that
is more than 3 kpc away. We see in this \figurename\ that the effect
of distance and realistic line-of-sight velocity uncertainties is
negligible. The exact shape and parameters of the stellar distribution
function also only minimally affect the predicted line-of-sight
velocity distribution. Changing the scale-length within its current
uncertainty changes the prediction only in the amplitude of the
effect: The Hercules stream is more prominent for shorter
scale-lengths since then there are relatively more stars in the inner
Galaxy that are brought to the Solar circle by the interaction with
the bar. Using a Shu distribution function \citep{shu69a} instead of
the Dehnen distribution function also gives very similar
predictions. Changing the axisymmetric potential by adjusting the
shape of the rotation curve shifts the peak as expected
\citep[see][]{dehnen00a}, but the Hercules feature nevertheless
remains detectable. The properties of the bar itself, \eg, its pattern
speed through \Rolr\ or its strength $\alpha$, change the predicted
pattern, but they do not significantly alter the prospects for
detection.

@@

The astrometric \Gaia\ mission will measure the parallaxes and proper
motions of up to one billion stars, most of which will be disk stars
\citep{Perryman01a}. With parallax and proper motion accuracies down
to 10 microarcseconds and mission-averaged line-of-sight velocity
uncertainties smaller than 10 km s$^{-1}$ for many stars to 17th
magnitude, \Gaia\ will be able to probe the kinematics of the disk out
to several kpc in all directions
\citep{Katz04a,bailerjones08a}. Crucially, it will provide large
samples of giants in many of the promising regions in
\figurename~\ref{fig:rphi2d}, with velocity measurements accurate
enough to verify the predictions laid out in this paper. This will
also allow the dynamical properties of the bar to be tightly
constrained, something that will be hard for \Gaia\ to do using direct
observations of the bulge due to the large extinction toward the
Galactic center \citep{robin05a}.

@@

\bibitem[Bailer-Jones(2008)]{bailerjones08a}
  Bailer-Jones, ~C.~A.~L., 2008,
  in IAU Symp.~254, ed.~J.~Andersen, J.~Bland-Hawthorn, and B.~Nordstr\"{o}m, (Dordrecht: Kluwer), 475

@@

\bibitem[Bovy, Hogg, \& Rix(2009)]{Bovy09b}
  Bovy,~J., Hogg,~D.~W., \& Rix, H.-W., 2009,
  \apj, 704, 1704

\bibitem[Bovy, Hogg, \& Roweis(2009)]{Bovy09a} Bovy,~J., Hogg,~D.~W., \& Roweis,~S.~T., 2009,
  \apj, 700, 1794

@@

@@

\bibitem[{{ESA}(1997)}]{ESA97a}
{ESA}, 1997, {The \emph{Hipparcos} and Tycho Catalogues} (ESA SP-1200; Noordwijk: ESA)

@@

\bibitem[Sellwood(2010)]{sellwood10a}
  Sellwood,~J.~A., 2010,
  \mnras, in press,
  arXiv:1001.5197v1 [astro-ph.GA]

@@

\bibitem[Shen \etal(2010)]{Shen10a}
  Shen,~J., \etal, 2010, \apj, 720L, 72

@@

\clearpage
\begin{figure}
\includegraphics[width=0.5\textwidth]{veldist.eps}
\caption{The velocity distribution of nearby ($\lesssim 100$ pc) stars
  from \citet{Bovy09a}. \vR\ and \vphi\ are the velocities toward the
  Galactic center and in the direction of Galactic rotation,
  respectively, corrected for the Solar motion. The Hercules moving
  group is visible as an overdensity at \vphi $\approx$ -0.2
  $\vo$.}\label{fig:obs}
\end{figure}

@@

\clearpage
\begin{figure}
\includegraphics[width=0.32\textwidth]{distuncertain.ps}
\includegraphics[width=0.32\textwidth]{vlosuncertain.ps}
\includegraphics[width=0.32\textwidth]{df.ps}\\
\includegraphics[width=0.32\textwidth]{slope.ps}
\includegraphics[width=0.32\textwidth]{rolr.ps}
\includegraphics[width=0.32\textwidth]{barstrength.ps}
\caption{Variation of the predicted velocity distribution at R = 1.075
  \Ro , $\phi = -21^{\circ}$ (d= 0.4 \Ro, $l$ = 270$^{\circ}$) with
  distance uncertainties; line-of-sight velocity uncertainties;
  parameters and shape of the distribution function; shape of the
  rotation curve; bar \Rolr; and bar strength.}\label{fig:1dvar}
\end{figure}

>>>>

In \figurename~\ref{fig:rphi1d} I show the predicted line-of-sight
velocity distributions in various locations near the Solar circle all
around the Galaxy, obtained by integrating the full two-dimensional
velocity distribution over the velocity component perpendicular to the
line of sight. This \figurename\ is no longer symmetric with respect
to shifts of 180$^{\circ}$ in azimuth, since the line-of-sight
direction breaks this symmetry. The gray curve is the line-of-sight
distribution we would observe if the targeted stars were in a steady
state in an axisymmetric potential with a Dehnen distribution function
(\eqnname~[\ref{eq:fdehnen}]). Both line-of-sight velocity
distributions are normalized to have equal mass since I assume that
our knowledge of the Galaxy and the tracer population will not be good
enough that we can test predicted differences in the absolute number
of stars at a given location between the axisymmetric and the
bar-resonance predictions. These differences are also largely a
function of the particular form and parameters of the distribution
function that we choose to represent the stellar disk before bar
formation.

@@

To further analyze the directions to target to best see the Hercules
feature in the line-of-sight velocities, I have calculated the
line-of-sight velocity distribution on a 100 by 20 grid in
Galactocentric azimuth and radius. As a measure of the difference
between the bar-resonance model and the steady-state, axisymmetric
model, I have calculated the Kullback-Leibler divergence between these
predicted distributions \citep[\eg,][]{mackay}
\begin{equation}
D_{\text{KL}}(P||Q) = \int \dd x\,p(x)\,\ln \frac{p(x)}{q(x)}\,,
\end{equation}
where, in this case, $x$ is the line-of-sight velocity, $p(x)$ is the
predicted distribution for the bar-origin model of the Hercules
stream, and $q(x)$ is the axisymmetric prediction. The
Kullback-Leibler divergence is only one among many possible choices to
quantify the difference between two probability distributions. It can
be interpreted as the expected weight of evidence for the prediction
by the bar-resonance model over the fully axisymmetric prediction that
can be expected from each sampled star. As such, locations with the
maximal Kullback-Leibler divergence are those in which the Hercules
feature stands out most prominently in the line of sight.

@@

Finally, in \figurename~\ref{fig:1dvar} I show the influence of
various nuisance and dynamical parameters on the predicted
line-of-sight velocity distribution for the most promising region that
is more than 3 kpc away. We see in this \figurename\ that the effect
of distance and realistic line-of-sight velocity uncertainties is
negligible. The exact shape and parameters of the stellar distribution
function also only minimally affect the predicted line-of-sight
velocity distribution. Changing the scale length within its current
uncertainty changes the prediction only in the amplitude of the
effect: The Hercules stream is more prominent for shorter scale
lengths since then there are relatively more stars in the inner Galaxy
that are brought to the Solar circle by the interaction with the
bar. Using a Shu distribution function \citep{shu69a} instead of the
Dehnen distribution function also gives very similar
predictions. Changing the axisymmetric potential by adjusting the
shape of the rotation curve shifts the peak as expected
\citep[see][]{dehnen00a}, but the Hercules feature nevertheless
remains detectable. The properties of the bar itself, \eg, its pattern
speed through \Rolr\ or its strength $\alpha$, change the predicted
pattern, but they do not significantly alter the prospects for
detection.

@@

The astrometric \Gaia\ mission will measure the parallaxes and proper
motions of up to one billion stars, most of which will be disk stars
\citep{Perryman01a}. With parallax and proper motion accuracies down
to 10 $\mu$as and mission-averaged line-of-sight velocity
uncertainties smaller than 10 km s$^{-1}$ for many stars to 17th
magnitude, \Gaia\ will be able to probe the kinematics of the disk out
to several kpc in all directions
\citep{Katz04a,bailerjones08a}. Crucially, it will provide large
samples of giants in many of the promising regions in
\figurename~\ref{fig:rphi2d}, with velocity measurements accurate
enough to verify the predictions laid out in this paper. This will
also allow the dynamical properties of the bar to be tightly
constrained, something that will be hard for \Gaia\ to do using direct
observations of the bulge due to the large extinction toward the
Galactic center \citep{robin05a}.

@@

\bibitem[Bailer-Jones(2008)]{bailerjones08a}
  Bailer-Jones, ~C.~A.~L., 2008,
  in IAU Symp.~254, The Galaxy Disk in Cosmological Context, ed.~J.~Andersen, J.~Bland-Hawthorn, \& B.~Nordstr\"{o}m, (Dordrecht: Kluwer), 475

@@

@@

\bibitem[Bovy \etal(2009a)]{Bovy09b}
  Bovy,~J., Hogg,~D.~W., \& Rix, H.-W., 2009a,
  \apj, 704, 1704

\bibitem[Bovy \etal(2009b)]{Bovy09a} Bovy,~J., Hogg,~D.~W., \& Roweis,~S.~T., 2009b,
  \apj, 700, 1794

@@

\bibitem[{{ESA}(1997)}]{ESA97a}
{ESA}, 1997, {The \emph{Hipparcos} and \emph{Tycho} Catalogues} (ESA SP-1200; Noordwijk: ESA)

@@

\bibitem[Sellwood(2010)]{sellwood10a}
  Sellwood,~J.~A., 2010,
  \mnras, 409, 145

@@

\bibitem[Shen \etal(2010)]{Shen10a}
  Shen,~J., \etal, 2010, \apj, 720, L72

@@

\clearpage
\begin{figure}
\includegraphics[width=0.5\textwidth]{veldist.eps}
\caption{Velocity distribution of nearby ($\lesssim 100$ pc) stars
  from \citet{Bovy09a}. \vR\ and \vphi\ are the velocities toward the
  Galactic center and in the direction of Galactic rotation,
  respectively, corrected for the Solar motion. The Hercules moving
  group is visible as an overdensity at \vphi $\approx$ -0.2
  $\vo$.}\label{fig:obs}
\end{figure}

@@

\clearpage
\begin{figure}
\includegraphics[width=0.32\textwidth]{distuncertain.ps}
\includegraphics[width=0.32\textwidth]{vlosuncertain.ps}
\includegraphics[width=0.32\textwidth]{df.ps}\\
\includegraphics[width=0.32\textwidth]{slope.ps}
\includegraphics[width=0.32\textwidth]{rolr.ps}
\includegraphics[width=0.32\textwidth]{barstrength.ps}
\caption{Variation of the predicted velocity distribution at $R$ =
  1.075 \Ro , $\phi = -21^{\circ}$ ($d$= 0.4 \Ro, $l$ = 270$^{\circ}$)
  with distance uncertainties; line-of-sight velocity uncertainties;
  parameters and shape of the distribution function; shape of the
  rotation curve; bar \Rolr; and bar strength.}\label{fig:1dvar}
\end{figure}
